---
title: "Exploratory Analysis - Stockholm Mortality Rate"
author: "Steve White"
date: "10/7/2021"
output:
  md_document:
    variant: gfm
always_allow_html: true
---

<style type="text/css">
  p,ul,li{
  font-size: 12pt;
}
</style>

```{r setup, include=FALSE}
library(htmltools)
library(tidyverse)
library(magrittr)
library(plotly)
library(tsibble)
library(feasts)
library(fable)

library(lubridate)
library(latex2exp)

knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE, results = "asis")
res_plots <- readRDS("data/res_plots.rds")
county_ts <- readRDS("data/county_ts.rds")

```

## Overview  

This document answers exercise questions from lesson 4 as well as explores the TS analysis framework provided by:
[https://otexts.com/fpp3/arima-r.html](https://otexts.com/fpp3/arima-r.html)


[Code Available on GitLab](https://gitlabstats-prd/StatsMethods/time-series-network/learning/arima-modelling/lecture-4/-/tree/swhite)  

## Analyse Raw Data {.tabset .tabset-pills}



```{r}

cat('\n') 
cat("### Observed data  \n")
htmltools::tagList(res_plots$stockholm_raw  )
cat('\n') 

# cat("### ACF  \n")
# btc_daily %>% ACF(close) %>% autoplot()
# cat('\n\n') 
# cat("### First Difference  \n")
# p <- btc_daily %>% autoplot(difference(close)) +
#   labs(title="Daily Close Price - 1st Diff",
#        y="USD") 
# htmltools::tagList(ggplotly(p))
# cat('\n\n') 
# cat("### ACF - First Diff  \n")
# btc_daily %>% ACF(difference(close)) %>% autoplot()
# 
# cat('\n\n')

  
```

## {-}



## Transformations {.tabset .tabset-pills}

The data is visually assessed to see if transformation or differencing is required.  Additionally a KPSS test is performed in which the null hypothesis is that the data is stationary (small p-value -> differencing is requried).

Logging the data appears to do little in altering the shape of the data.  Differencing appears to make the data more stationary.    Also KPSS p-value of the orignal series is reported as .01 which indicates to reject the null.  The KPSS p-value for the diffed series is reported as 0.1.  Therefore diffed data will be used for model selection. 

### actual-data

```{r, results = "hold"}

# cat('\n') 
# cat("### actual-data  \n")
county_ts %>%  gg_tsdisplay(mort_rate, plot_type='partial', lag_max = 12)
county_ts %>% features(mort_rate, unitroot_kpss)

```

### log-data 

```{r, results = "hold"}

# cat('\n\n') 
# cat("### log-data  \n")
county_ts %>%  gg_tsdisplay(log(mort_rate), plot_type='partial', lag_max = 12)
county_ts %>% features(log(mort_rate), unitroot_kpss)
```

### diff-data 

```{r, results = "hold"}

# cat('\n\n') 
# cat("### diff-data  \n")
county_ts %>%  gg_tsdisplay(difference(mort_rate), plot_type='partial', lag_max = 12)
county_ts %>% features(difference(mort_rate), unitroot_kpss)
# cat('\n\n') 

```

### logdiff-data

```{r, results = "hold"}
# cat("### logdiff-data  \n")
county_ts %>%  gg_tsdisplay(difference(log(mort_rate)), plot_type='partial', lag_max = 12)
county_ts %>% features(difference(log(mort_rate)), unitroot_kpss)
# cat('\n\n') 


```

## {-}



## Model Selection {.tabset .tabset-pills}

```{r, results = "hold"}
county_model <- county_ts  %>% model(
  arima011 = ARIMA(mort_rate ~ pdq(0,1,1)),
  stepwise = ARIMA(mort_rate)
  )
```

The ACF appears to be significant at lag 1 and then cut off.  The PACF appears to show geometric decay.  I will choose an MA(1) model for the diffed series or ARIMA(0,1,1) on the original (ie the airline model). The automatic model was performed using the stepwise [Hyndman-Khandakar algorithm](https://otexts.com/fpp3/arima-r.html).  The algorithm chose an ARIMA(2,1,0) Which I would have never chosen by looking at the ACF and PACF.


Summary of manual model and automatic model:
```{r, results = "hold"}
glance(county_model)
```


### Manual

```{r, results = "hold"}

county_model %>% select(arima011) %>%report()
county_model %>% select(arima011) %>% gg_tsresiduals()


```



### Auto  


```{r, results = "hold"}

county_model %>% select(stepwise) %>%report()
county_model %>% select(stepwise) %>% gg_tsresiduals()


```


## {-}

## Forecast {.tabset .tabset-pills}
### Manual

```{r, results = "hold"}
county_model %>%
  forecast(h=5)  %>% filter(.model=='arima011') %>%
  autoplot(county_ts)
```

### Auto

```{r, results = "hold"}
county_model %>%
  forecast(h=5)  %>% filter(.model=='stepwise') %>%
  autoplot(county_ts)
```

## {-}



